---
title: Introduction
output: html_document
---

## Overview

*EvoTraceR* is an R package to identifiy and analyse Amplicon Sequence Variants after performing an experiment through the EVoBC kit. 

```{r include=FALSE}
library(knitr)
opts_chunk$set(
concordance = TRUE,
background = "#f3f3ff"
)
```

## Installing EvoTraceR

The R version of *EvoTraceR* can be installed from Github. To do so, we need to install the R packages *EvoTraceR* depends on and the devtools package. 


```{r eval=FALSE}
# install EvoTraceR dependencies
if (!require("devtools")) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("cli")) install.packages("cli")
if (!require("Biostrings")) BiocManager::install("Biostrings")
if (!require("benthos")) install.packages("benthos")
#if (!require("lemon")) install.packages("lemon")
if (!require("scales")) install.packages("scales")
if (!require("utils")) install.packages("utils")
if (!require("methods")) install.packages("methods")
if (!require("data.table")) install.packages("data.table")
if (!require("ggtree")) BiocManager::install("ggtree")
if (!require('foreach')) install.packages('foreach')
if (!require('reticulate')) install.packages('reticulate')

Users should also have a virtual enviroment (created for example thorugh anaconda), where the python dependencies are installed.
In case users want this to be taken care of automatically, they can unload reticulate (in case it was already loaded) prior to installing EvoTraceR. Otherwise, they can setup an anaconda enviroment where all dependencies ahve been manually installed.
The dependencies are the following:
Numpy version 1.21.5
Numba
Cassiopeia-lineage
umi_tools
```{r eval=FALSE}
# Run only if reticulate has been manually setup:
# library(reticulate)
# reticulate::use_condaenv('EvoTraceR_enviroment')
# Install EvoTraceR
devtools::install_github("Nowak-Lab/EvoTraceR", ref = "master")
```


In order to handle fastq files with forward and reverse reads, EvoTraceR uses trimmomatic and flash tools to perform adapters trimming, discard low quality bases and merge
forward and reverse reads. In order to use such packages, users should install [trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) and [flash]{https://ccb.jhu.edu/software/FLASH/}.
```{r eval=FALSE}
# load EvoTraceR library
library("EvoTraceR")
```

Finally, in case reticulate is setpup automatically through the install process, users need to call the following function to finish setup and then restart R. 
```{r eval=FALSE}
setup_reticulate()
```

## Debug

Please feel free to contact us if you have problems running our tool at l.patruno@campus.unimib.it. 
