---
title: Introduction
output: html_document
---

## Overview

*EvoTraceR* is an R package to identifiy and analyse Amplicon Sequence Variants after performing an experiment through the EVoBC kit. 

```{r include=FALSE}
library(knitr)
opts_chunk$set(
concordance = TRUE,
background = "#f3f3ff"
)
```

## Installing EvoTraceR

The R version of *EvoTraceR* can be installed from Github. To do so, we need to install the R packages *EvoTraceR* depends on and the devtools package. 


```{r eval=FALSE}
# install EvoTraceR dependencies
if (!require("devtools")) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("cli")) install.packages("cli")
if (!require("Biostrings")) BiocManager::install("Biostrings")
if (!require("benthos")) install.packages("benthos")
if (!require("lemon")) install.packages("lemon")
if (!require("scales")) install.packages("scales")
# if (!require("ggmsa")) devtools::install_github("YuLab-SMU/ggmsa")
if (!require("utils")) install.packages("utils")
if (!require("methods")) install.packages("methods")
if (!require("data.table")) install.packages("data.table")
if (!require("ggtree")) BiocManager::install("ggtree")
if (!require("TreeTools")) install.packages('TreeTools')
if (!require('ade4')) install.packages('ade4')
if (!require('foreach')) install.packages('foreach')
if (!require('reticulate')) install.packages('reticulate')

# install devtools library and EvoTraceR

devtools::install_github("Nowak-Lab/EvoTraceR", ref = "master")
```

In order to handle fastq files with forward and reverse reads, EvoTraceR uses trimmomatic and flash tools to perform adapters trimming, discard low quality bases and merge
forward and reverse reads. In order to use such packages, users should install [trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) and [flash]{https://ccb.jhu.edu/software/FLASH/}.

```{r eval=FALSE}
# load EvoTraceR library
library("EvoTraceR")
```

## Debug

Please feel free to contact us if you have problems running our tool at l.patruno@campus.unimib.it. 
