---
title: Introduction
output: html_document
---

## Overview

*REvoBC* is an R package to identifiy and analyse Amplicon Sequence Variants after performing an experiment through the EVoBC kit. 

```{r include=FALSE}
library(knitr)
opts_chunk$set(
concordance = TRUE,
background = "#f3f3ff"
)
```

## Installing REvoBC

The R version of *REvoBC* can be installed from Github. To do so, we need to install the R packages *REvoBC* depends on and the devtools package. 


```{r eval=FALSE}
# install REvoBC dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require("dada2")) BiocManager::install("dada2")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("cli")) install.packages("cli")
if (!require("Biostrings")) BiocManager::install("Biostrings")
if (!require("benthos")) install.packages("benthos")
if (!require("lemon")) install.packages("lemon")
if (!require("scales")) install.packages("scales")
if (!require("muscle")) BiocManager::install("muscle")
if (!require("R4RNA")) BiocManager::install("R4RNA")
if (!require("ggmsa")) install.packages("ggmsa")
if (!require("utils")) install.packages("utils")
if (!require("methods")) install.packages("methods")
if (!require("data.table")) install.packages("data.table")

# install devtools library and REvoBC
if (!require("devtools")) install.packages("devtools")
library("devtools")
install_github("Nowak-Lab/REvoBC", ref = "master")

# load REvoBC library
library("REvoBC")
```

## Debug

Please feel free to contact us if you have problems running our tool at l.patruno@campus.unimib.it. 
