---
title: Introduction
output: html_document
---

## Overview

*EvoTraceR* is an R package to identifiy and analyse Amplicon Sequence Variants after performing an experiment through the EVoBC kit. 

```{r include=FALSE}
library(knitr)
opts_chunk$set(
concordance = TRUE,
background = "#f3f3ff"
)
```

## Installing EvoTraceR

The R version of *EvoTraceR* can be installed from Github. To do so, we need to install the R packages *EvoTraceR* depends on and the devtools package. 


```{r eval=FALSE}
# install EvoTraceR dependencies
if (!require("devtools")) install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("cli")) install.packages("cli")
if (!require("Biostrings")) BiocManager::install("Biostrings")
if (!require("benthos")) install.packages("benthos")
if (!require("lemon")) install.packages("lemon")
if (!require("scales")) install.packages("scales")
if (!require("phytools")) install.packages("phytools")
if (!require("muscle")) BiocManager::install("muscle")
if (!require("R4RNA")) BiocManager::install("R4RNA")
# if (!require("ggmsa")) devtools::install_github("YuLab-SMU/ggmsa")
if (!require("utils")) install.packages("utils")
if (!require("methods")) install.packages("methods")
if (!require("data.table")) install.packages("data.table")
if (!require("Rphylip")) BiocManager::install("Rphylip")
if (!require("phytols")) install.packages("phytools")
if (!require("ggtree")) BiocManager::install("ggtree")
if (!require("TreeTools")) install.packages('TreeTools')
if (!require("phylogram")) install.packages('phylogram')
if (!require('ade4')) install.packages('ade4')
if (!require('foreach')) install.packages('foreach')

# install devtools library and EvoTraceR

devtools::install_github("Nowak-Lab/EvoTraceR", ref = "master")
```

In order to build the phylogenetic tree, EvoTraceR uses the camin-sokal implementation found in the package Rphylip. In order to use such package, users should also install [mix](https://evolution.genetics.washington.edu/phylip.html).
The path to the mix binary needs to be provided to the EvoTraceR function that performs the phylogeny inference.

```{r eval=FALSE}
# load EvoTraceR library
library("EvoTraceR")
```

## Debug

Please feel free to contact us if you have problems running our tool at l.patruno@campus.unimib.it. 
