---
title: Running EvoTraceR
output: html_document
---

## Setup the environement

*EvoTraceR* is the computational tool inside the project EboBC. After performing EvoBC experiments in the lab, this tool can be exploited to perform the analysis of Amplicon Sequence Variants. 
It uses the algorithm dada2 to identify all ASVs, then performs multi-sequence alignment to identify mutations and finally performs the phylogeny reconstruction.
In order to perform the analysis, there are multiple functions that perform the different steps of the analysis pipeline. 
The first step initializes the EvoTraceR object, which is a list that gets populated step-by-step with multiple fields containing the result of the analysis. 


```{r}
library("EvoTraceR")
```

The initialization function computes the set of Amplicon Sequence Variants.
It calls trimmomatic and flash tools to perform adapters trimming, discard low quality bases and merge forward and reverse reads.
The main parameters for the initialization are the directory where the output will be stored and one of the following: 
input directory: in this case users have the fastq files and want to perform the whole analysis using EvoTraceR. In this case they need to provide the path to the folder containing the fastq (which may be compressed), and then they can provide any additional parameter for dada2 functions. 
trimmomatic_path: Local path to the executable of Trimmomatic. 
flash_path. Local path to the executable of \code{Flash}.
Important: each fastq file pair (forward and reverse) must be associated to the sample on which the experiment was performed. The sample can be either the organ of the moouse or the an ID for the timepoint in case longitudinal data is provided. 
In order for EvoTraceR to associate each fastq pair to its sample users have two options: 
one is to name the fastq following this pattern: FILEPREFIX_SAMPLE_FILESUFFIX_RX.fastq, where FILEPREFIX and FILESUFFIX can be any string (without the "_" character). In this case EvoTraceR will consider as sample ID the second field separated by the underscore. 
The second option is to provide a list that associates each filename (without the suffix _R1 and _R2) to the corresponding sample. For example, if we have forward and reverse files named file1_R1.fastq and file1_R2.fastq that correspond to organ PRL (code for Prostate Left), than the parameter should be set as: map_file_sample = c("file1" = "PRL")). 

```{r}
input_dir = system.file("extdata", "input", package = "EvoTraceR")
output_dir = system.file("extdata", "output", package = "EvoTraceR")
```

We can now perform the initialization. 

```{r}
EvoTraceR_object = initialize_EvoTraceR(input_dir = input_dir, 
                                        output_dir = output_dir, 
                                        trimmomatic_path = '~/Trimmomatic 0.xx/trimmomatic-0.xx.jar',
                                        flash_path = '~/*/FLASH-x.x.x/flash')
names(EvoTraceR_object)
```

During the above computation, all unique sequences from each sample have been identify. In the next steps these sequences will be cleaned.

## Analysis and pairwise alignment of ASVs

This function performs the analysis on ASV sequences identified by the previous steps and aligns them to the reference sequence.
First, it pools together those sequences characterized by a Hamming distance equals or lower than 2. 
Then it performs pairwise alignment using Needleman-Wunsch global alignment algorithm implemented in function \code{pairwiseAlignment}
in package \code{Biostrings}, aligning each sequence to the original barcode considered in the analysis 
(See the Biostrings documentation \href{https://www.rdocumentation.org/packages/Biostrings/versions/2.40.2/topics/pairwiseAlignment}{here} for more details).
Next it pools together those sequences that differ by one another only by substitutions, summing their counts, and after this it removes
those sequences showing a frequency lower than the threshold specified in parameter \code{asv_count_cutoff}
Finally it performs filtering of the sequences based on their flanking sequances.
Then, it computes different statistics for the final ASVS, storing:
the relative frequency of all ASVs in each sample. 
the relative frequency of each ASV in the samples.
the counts for each ASV normalized to the counts of the sample with maximum frequency
the frequency of the different ASVs in each sample.

```{r}
EvoTraceR_object = asv_analysis(EvoTraceR_object = EvoTraceR_object)
names(EvoTraceR_object)
```

There are new fields: 
clean_asv_dataframe which contains the counts detected for each ASV after filtering (i.e. after filtering for contamination identification of NMBC). 
reference is the information on the reference sequence considered for the analysis.
clean_asv_dataframe: ASV sequences identified post-filtering (contamination removed,
sequences with a similarity higher than \code{pid_cutoff_nmbc} to the original barcode
aggregated to it and ASVs named in increasing order (ASV01, ASV02, etc.) according
to their total counts. 
reference: Info about the reference sequence used for the current analysis. 
statistics: is a new list that contains all the results of the computation (see the manual for a detailed explanation). Among its fileds there are asv_df_percentages, which contains the normalized counts, and asv_diversity_perSample which contains the measures of clonal heterpgeneity. 
alignemnt: is a new list containing the result of the alignment step. It contains two fields:  msa_stringset is the output of multiple sequence alignemnt performed by muscle; asv_barcode_alignment is a tibble
where each line corresponds to a nucleotide in each ASV with the corresponding sample in which it was found, and it indicates whether the nucletide is mutated and the type of alteration.

```{r}
EvoTraceR_object = analyse_mutations(EvoTraceR_object)
```

After alignment, EvoTraceR identifies, for each position in the original reference, the number of alterations found on the different ASVs. The final result is shown in figure gghist_del_sub_ins_perc.pdf which indicates for each position the frequency of the different mutations in each sample.
This computation stores the result in a new field mutations_df. 

```{r eval=FALSE}
EvoTraceR_object = infer_phylogeny(EvoTraceR_object)
```
This function reconstructs the phylogenetic tree using the Cassiopeia's suite. 
It uses the greedy algorithm, that iteratively splits the set of ASVs in clusters
based on the most common mutation at each iteration.

```{r eval=FALSE}
EvoTraceR_object = create_df_summary(EvoTraceR_object)
```

Finally, this function combines in one dataframe all the data that have been computed for the phylogenetic analysis.